rgbif: R client for working with GBIF species occurrence data

### 2015-10-28 manuscript.Rmd
updated names section

`rgbif` is designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, species names,
occurrences, and maps. We ignore maps in `rgbif` as it is concerned with
generating maps primarily for web applications. `rgbif` has a suite of functions dealing
with each of registry, species names, and occurrences - we'll go through
each in turn describing design and example usage.

@@

```{r}
res <- dataset_metrics(uuid='ec93a739-1681-4b04-b62f-3a687127a17f')
df <- data.frame(rank = names(res$countByRank),
                 count = unname(unlist(res$countByRank)))
knitr::kable(df)
```

@@

## Species

@@

* `occ_count()`
* `occ_search()`
* `occ_get()`
* `occ_metadata()`

>>>>

`rgbif` is designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, taxonomic names,
occurrences, and maps. We ignore maps in `rgbif` as it is concerned with
generating maps primarily for web applications. `rgbif` has a suite of functions dealing
with each of registry, taxonomic names, and occurrences - we'll go through
each in turn describing design and example usage.

@@

```{r eval=FALSE}
res <- dataset_metrics(uuid='ec93a739-1681-4b04-b62f-3a687127a17f')
data.frame(rank = names(res$countByRank),
           count = unname(unlist(res$countByRank)))
```

```{r echo=FALSE}
res <- dataset_metrics(uuid='ec93a739-1681-4b04-b62f-3a687127a17f')
df <- data.frame(rank = names(res$countByRank),
                 count = unname(unlist(res$countByRank)))
knitr::kable(df)
```

@@

## Taxonomic names

The GBIF taxonomic names API services are spread across five functions:

* Search GBIF name backbone - `name_backbone()`
* Search across all checklists - `name_lookup()`
* Quick name lookup - `name_suggest()`
* Name usage of a name according to a checklist - `name_usage()`
* GBIF name parser - `parsenames()`

The goal of these name functions is often to settle on a taxonomic name
known to GBIF's database. This serves two purposes: 1) when referring to
a taxonomic name, you can point to a URI on the internet, and 2) you can
search for metadata on that species, and occurrences of that species in
GBIF.

Taxonomic names are particularly tricky. Many different organizations have
their own unique codes for the same taxonomic names, and some taxonomic
groups have preferred sources for the definitive names for that group.

When searching for occurrences (see below) you can search by taxonomic
name (and other filters, e.g., taxonomic rank), but you're probably better off
figuring out the taxonomic key in the GBIF backbone taxonomy, and using
that to search for occurrences. The `taxonkey` parameter in the GBIF occurrences
API expects a GBIF backbone taxon key.

### GBIF Backbone

The GBIF backbone taxonomy is used in GBIF to have a consistent way to refer
to taxonomic names throughout their services. The backbone has `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$distinctNamesCount` unique
names and `r dataset_metrics("d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")$countByRank$SPECIES` species names. The backbone taxonomy is also a dataset with key
`d7dddbf4-2cf0-4f39-9b2a-bb099caae36c`
(http://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c).

We can search the backbone taxonomy with the function `name_backbone()`. Here,
we're searching for the name _Poa_, restricting to genera, and the family
_Poaceae_.

```{r}
res <- name_backbone(name='Poa', rank='genus', family='Poaceae')
res[c('usageKey', 'kingdom')]
```

### Name searching

One of the quickest ways to search for names is using `name_suggest()`, which
does a very quick search and returns minimal data. Here, we're searching for
the query tem _Pum_, and we get back a bunch of names:

```{r eval=FALSE}
name_suggest(q='Pum', limit = 6)
```

```{r echo=FALSE}
df <- name_suggest(q='Pum', limit = 6)
knitr::kable(df)
```

@@

* Get a single numeric count of occurrenes - `occ_count()`
* Search for occurrences - `occ_search()`
* Get occurrences by occurrence identifier - `occ_get()`
* Get occurrence metadata - `occ_metadata()`

### 2015-10-28 manuscript.tex
updated names section

\texttt{rgbif} is designed following the
\href{http://www.gbif.org/developer/summary}{GBIF Application
Programming Interface}, or API. The GBIF API has four major components:
registry, species names, occurrences, and maps. We ignore maps in
\texttt{rgbif} as it is concerned with generating maps primarily for web
applications. \texttt{rgbif} has a suite of functions dealing with each
of registry, species names, and occurrences - we'll go through each in
turn describing design and example usage.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_metrics}\NormalTok{(}\DataTypeTok{uuid=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\NormalTok{df <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{rank =} \KeywordTok{names}\NormalTok{(res$countByRank),}
                 \DataTypeTok{count =} \KeywordTok{unname}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(res$countByRank)))}
\NormalTok{knitr::}\KeywordTok{kable}\NormalTok{(df)}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Species}\label{species}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"dplyr"}\NormalTok{)}
\NormalTok{dat <-}\StringTok{ }\KeywordTok{occ_download_import}\NormalTok{(res)}
\NormalTok{dat %>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(gbifID, decimalLatitude, decimalLongitude)}
\CommentTok{#>       gbifID decimalLatitude decimalLongitude}
\CommentTok{#> 1  657590544              NA               NA}
\CommentTok{#> 2  657679551              NA               NA}
\CommentTok{#> 3  657791316        37.70805        -118.4162}
\CommentTok{#> 4  658180562              NA               NA}
\CommentTok{#> 5  441881672              NA               NA}
\CommentTok{#> 6  911596181              NA               NA}
\CommentTok{#> 7   56454601              NA               NA}
\CommentTok{#> 8  657848913              NA               NA}
\CommentTok{#> 9  658187373              NA               NA}
\CommentTok{#> 10 658279212        38.95917        -106.9892}
\CommentTok{#> ..       ...             ...              ...}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \texttt{occ\_count()}
\item
  \texttt{occ\_search()}
\item
  \texttt{occ\_get()}
\item
  \texttt{occ\_metadata()}
\end{itemize}

>>>>

\texttt{rgbif} is designed following the
\href{http://www.gbif.org/developer/summary}{GBIF Application
Programming Interface}, or API. The GBIF API has four major components:
registry, taxonomic names, occurrences, and maps. We ignore maps in
\texttt{rgbif} as it is concerned with generating maps primarily for web
applications. \texttt{rgbif} has a suite of functions dealing with each
of registry, taxonomic names, and occurrences - we'll go through each in
turn describing design and example usage.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_metrics}\NormalTok{(}\DataTypeTok{uuid=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{rank =} \KeywordTok{names}\NormalTok{(res$countByRank),}
           \DataTypeTok{count =} \KeywordTok{unname}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(res$countByRank)))}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Taxonomic names}\label{taxonomic-names}

The GBIF taxonomic names API services are spread across five functions:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Search GBIF name backbone - \texttt{name\_backbone()}
\item
  Search across all checklists - \texttt{name\_lookup()}
\item
  Quick name lookup - \texttt{name\_suggest()}
\item
  Name usage of a name according to a checklist - \texttt{name\_usage()}
\item
  GBIF name parser - \texttt{parsenames()}
\end{itemize}

The goal of these name functions is often to settle on a taxonomic name
known to GBIF's database. This serves two purposes: 1) when referring to
a taxonomic name, you can point to a URI on the internet, and 2) you can
search for metadata on that species, and occurrences of that species in
GBIF.

Taxonomic names are particularly tricky. Many different organizations
have their own unique codes for the same taxonomic names, and some
taxonomic groups have preferred sources for the definitive names for
that group.

When searching for occurrences (see below) you can search by taxonomic
name (and other filters, e.g., taxonomic rank), but you're probably
better off figuring out the taxonomic key in the GBIF backbone taxonomy,
and using that to search for occurrences. The \texttt{taxonkey}
parameter in the GBIF occurrences API expects a GBIF backbone taxon key.

\subsubsection{GBIF Backbone}\label{gbif-backbone}

The GBIF backbone taxonomy is used in GBIF to have a consistent way to
refer to taxonomic names throughout their services. The backbone has
4410899 unique names and 2497114 species names. The backbone taxonomy is
also a dataset with key \texttt{d7dddbf4-2cf0-4f39-9b2a-bb099caae36c}
(\url{http://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c}).

We can search the backbone taxonomy with the function
\texttt{name\_backbone()}. Here, we're searching for the name
\emph{Poa}, restricting to genera, and the family \emph{Poaceae}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{name_backbone}\NormalTok{(}\DataTypeTok{name=}\StringTok{'Poa'}\NormalTok{, }\DataTypeTok{rank=}\StringTok{'genus'}\NormalTok{, }\DataTypeTok{family=}\StringTok{'Poaceae'}\NormalTok{)}
\NormalTok{res[}\KeywordTok{c}\NormalTok{(}\StringTok{'usageKey'}\NormalTok{, }\StringTok{'kingdom'}\NormalTok{)]}
\CommentTok{#> $usageKey}
\CommentTok{#> [1] 2704173}
\CommentTok{#> }
\CommentTok{#> $kingdom}
\CommentTok{#> [1] "Plantae"}
\end{Highlighting}
\end{Shaded}

\subsubsection{Name searching}\label{name-searching}

One of the quickest ways to search for names is using
\texttt{name\_suggest()}, which does a very quick search and returns
minimal data. Here, we're searching for the query tem \emph{Pum}, and we
get back a bunch of names:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{name_suggest}\NormalTok{(}\DataTypeTok{q=}\StringTok{'Pum'}\NormalTok{, }\DataTypeTok{limit =} \DecValTok{6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}rll@{}}
\toprule
key & canonicalName & rank\tabularnewline
\midrule
\endhead
3269133 & Pumilus & GENUS\tabularnewline
4407849 & Pumilinura & GENUS\tabularnewline
4323990 & Pumiliopsis & GENUS\tabularnewline
4324083 & Pumiliopes & GENUS\tabularnewline
4161281 & Pumilea & GENUS\tabularnewline
4312370 & Pumilopagurus & GENUS\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"dplyr"}\NormalTok{)}
\NormalTok{dat <-}\StringTok{ }\KeywordTok{occ_download_import}\NormalTok{(res)}
\NormalTok{dat %>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(gbifID, decimalLatitude, decimalLongitude)}
\CommentTok{#>       gbifID decimalLatitude decimalLongitude}
\CommentTok{#> 1  725767384        61.01005         24.41740}
\CommentTok{#> 2  725767447        59.82923         23.13550}
\CommentTok{#> 3  725767450        60.38505         25.17449}
\CommentTok{#> 4  725767513        68.37648         23.51963}
\CommentTok{#> 5  725767546        67.19203         24.85820}
\CommentTok{#> 6  725767579        60.21607         24.67412}
\CommentTok{#> 7  725767609        66.49260         25.70471}
\CommentTok{#> 8  725767645        61.36634         24.76218}
\CommentTok{#> 9  725767678        62.29174         27.96500}
\CommentTok{#> 10 725767681        60.28615         22.38489}
\CommentTok{#> ..       ...             ...              ...}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Get a single numeric count of occurrenes - \texttt{occ\_count()}
\item
  Search for occurrences - \texttt{occ\_search()}
\item
  Get occurrences by occurrence identifier - \texttt{occ\_get()}
\item
  Get occurrence metadata - \texttt{occ\_metadata()}
\end{itemize}
